---
title: "Chronic granulomatous disease (CGD) analysis"
# output: rmarkdown::html_vignette
# output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Chronic granulomatous disease (CGD) analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, warning=FALSE, message=FALSE}
library(purrr)
library(reshape2)
library(dplyr)
library(rstan)
library(dplyr)
library(glue)
library(abind)
library(survival)
library(multimcm)
library(frailtyHL)

library(multimcm)

# rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores() - 1)
```

# Introduction

The analysis is a comparison with a frequentist analysis in @Lai2009.


# Data

The data is chronic granulomatous disease (CGD) data available as part of the `{frailtyHL}` package from @fleming1991. From the package help documents

> The CGD data set in Fleming and Harrington (1991) is from a placebo-controlled randomized trial of gamma interferon in chronic granulomatous disease. In total, 128 patients from 13 hospitals were followed for about 1 year. The number of patients per hospital ranged from 4 to 26. Each patient may experience more than one infection. The survival times (times-to-event) are the times between recurrent CGD infections on each patient (i.e. gap times). Censoring occurred at the last observation for all patients, except one, who experienced a serious infection on the date he left the study.


```{r}
data(cgd)
head(cgd)
```

Create a new variable for the duration until the event.
```{r}
cgd <- mutate(cgd, time = tstop - tstart)
# center_id = as.numeric(as.factor(center))
```

The data set has repeat measurements for recurrent infections which we currently have not implemented in `{multimcm}` so for simplicity we'll remove some of the data in the initial analyses.
In @Lai2009 hospital-level random effects are assigned to the cured probability part, and both patient- and hospital-level random effects are assigned to hazard function part. `{multimcm}` currently only applies hierarchical structure to the cure fraction and not to the separate hazards.

Take the first event for each individual by start time.
```{r first-measurement}
cgd_first <-
  cgd |> 
  group_by(id) |> 
  arrange(tstart) |> 
  filter(row_number() == 1)
```

The full data set also has 13 hospitals/centers so select a smaller number:

| Center               | Country        |
|----------------------|----------------|
| Amsterdam            | Netherlands    |
| NIH                  | United States  |
| University of Zurich | Switzerland    |
| Scripps Institute    | United States  |

Drop any levels which only appeared in the data we've just removed and convert the variables to numeric or factor for unique and categorical data, respectively.

```{r filter-center}
cgd_center <- cgd_first |>
  filter(center %in% c("Amsterdam", "NIH", "Univ. of Zurich", "Scripps Institute")) |>
  droplevels() |> 
  mutate(center_id = as.numeric(center),
         hos.cat_id = as.numeric(hos.cat),
         sex_id = as.factor(as.numeric(sex)))
```

Next, append the background hazard rate by age, sex and country. These rare are available from the [WHO website](https://www.who.int/data/gho/data/themes/topics/topic-details/GHO/healthy-life-expectancy-(hale)) @wholifetables. We will read from our own local package `{bgfscure}` for simplicity.
Next, we need to harmonise the country names and their associated cities.
Then we join the data with the background mortality data according to age, sex and country.
```{r bg-hazard}
# WHO background mortality data
load(here::here("../bgfscure/data/bg.mortality.RData"))  # bg.mortality

cgd_center <-
  cgd_center |>
  tidyr::separate(hos.cat, c("country", "hospital"),
                  remove = FALSE) |> 
  mutate(country = toupper(country),
         age_event = round(age + time/365.25, 0),   # convert from days to years
         country = ifelse(center == "Amsterdam", "NETHERLANDS",
                          ifelse(center == "Univ. of Zurich", "SWITZERLAND",
                                 ifelse(center == "Copenhagen", "DENMARK", country)))) |> 
  select(-hospital)

# harmonise fields
bg.mortality <- bg.mortality |> 
  mutate(ACOUNTRY = toupper(ACOUNTRY),
         ACOUNTRY = ifelse(ACOUNTRY == "UNITED STATES", "US", ACOUNTRY),
         SEX = ifelse(SEX == "M", "male", "female"),
         rate = ifelse(rate == 0, 1e-10, rate)) |>  # replace so >0
  rename(country = ACOUNTRY,
         sex = SEX,
         age_event = Age)

input_data <- merge(cgd_center, bg.mortality,
                    by = c("age_event", "sex", "country"), sort = FALSE)
```

Creating `input_data` is the last step in the data wrangling. We can now proceed to the mixture cure modelling analysis.


# Analysis

### By center

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda)\\
\pi_i = \text{logit}^{-1}(\alpha + \beta_{\text{treat}[i]} + \gamma_{\text{center}[i]})\\
\gamma_{\text{center}[i]} \sim N(\mu_{\text{center}}, \sigma^2_{\text{center}})
$$

```{r}
out <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400)
```

Alternatively, we can pass in a _precompiled_ model, rather than have the model compiled at the time of fitting.

```{r}
model_nm <- "precompiled_stan_model"

# create the precompiled model and save as .RDS file
precompile_bmcm_model(input_data = input_data,
                      cureformula =  "~ treat + (1 | center_id)",
                      family_latent = "exponential",
                      model_name = model_nm)
```

Now, load back in the compiled model and fit as before.

```{r}
model_path <- glue::glue("{system.file('stan', package = 'multimcm')}/{model_nm}.RDS")

out_precompiled <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path)
```

We can do the same but using a `cmdstanr` model, rather than an `rstan` model. This has some benefits in terms of speed, reliability and memory usage.
Firstly for a non-compiled model, set `use_cmdstanr = TRUE` argument.

```{r}
out <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    use_cmdstanr = TRUE)

# so that plotting functions work later
# convert cmdstanr object to rstan stanfit format

##TODO: error. for some reason save_warmup is NA?...
# out$output <- rstan::read_stan_csv(out$output$output_files())
```

Secondly, we can pass in a _precompiled_ `cmdstanr` model. This saves an `.exe` file in the `stan` folder.

```{r}
model_nm <- "precompiled_stan_model"

precompile_bmcm_model(input_data = input_data,
                      cureformula =  "~ treat + (1 | center_id)",
                      family_latent = "exponential",
                      model_name = model_nm,
                      use_cmdstanr = TRUE)
```

and load back in the compiled model and fit as before.

```{r}
model_path <- glue::glue("{system.file('stan', package = 'multimcm')}/{model_nm}.exe")

out_precompiled <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path,
    use_cmdstanr = TRUE)

##TODO:
# out_precompiled$output <-
#   rstan::read_stan_csv(out_precompiled$output$output_files())
```

#### Separate models

We can also fit the end-points separately by changing the `cureformula` argument to a formula without a random effect term `~ treat + center_id`.

```{r}
stan_model <- 
  precompile_bmcm_model(input_data = input_data,
                        cureformula =  "~ treat + center_id",
                        family_latent = "exponential",
                        model_name = "precompiled_stan_model",
                        use_cmdstanr = TRUE)

model_path <- stan_model$exe_file()

out_precompiled_separate <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + center_id",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path,
    use_cmdstanr = TRUE)
```

### By hospital category

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda)\\
\pi_i = \text{logit}^{-1}(\alpha + \beta_{\text{treat}[i]} + \gamma_{\text{cat}[i]})\\
\gamma_{\text{cat}[i]} \sim N(\mu_{\text{cat}}, \sigma^2_{\text{cat}})
$$

```{r}
out_hos.cat <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | hos.cat_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 365)
```


### With sex covariate in latent model

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda_i)\\
\log(\lambda_i) = \alpha_{\lambda} + \beta_{\text{sex}[i]}\\
\pi_i = \text{logit}^{-1}(\alpha_{\pi} + \beta_{\text{treat}[i]} + \gamma_{\text{center}[i]})\\
\gamma_{\text{center}[i]} \sim N(\mu_{\text{center}}, \sigma^2_{\text{center}})
$$

```{r}
out_sex <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ sex",
    cureformula = "~ treat + (1 | hos.cat_id)",
    family_latent = "exponential",
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 365)
```


## Plots

After fitting the models, we can plot the survival curves for each.

```{r}
library(ggplot2)
```

### Main models

Hierarchy by `center_id`.

```{r}
gg <- plot_S_joint(out,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

Hierarchy by `center_id` with precompiled stan model. Should be the same!

```{r eval=FALSE, echo=FALSE}
gg <- plot_S_joint(out_precompiled,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

Separate models, i.e fixed effect for `center_id`.
TODO: THIS DOESNT LOOK RIGHT


```{r eval=FALSE, echo=FALSE}
gg <- plot_S_joint(out_precompiled_separate,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg, filename = "plots/cgd_surv_plot.png", dpi = 640, width = 10, height = 8)
```

### By center

```{r}
gg2 <- plot_S_joint(out_hos.cat,
                    add_km = TRUE,
                    annot_cf = FALSE)
gg2 + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg2, filename = "plots/cgd_hospcat_surv_plot.png", dpi = 640, width = 10, height = 8)
```

```{r}
gg3 <- plot_S_joint(out_sex,
                    add_km = TRUE,
                    annot_cf = FALSE)
gg3 + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg3, filename = "plots/cgd_sex_surv_plot.png", dpi = 640, width = 10, height = 8)
```


## References
