---
title: "CGD analysis"
# output: rmarkdown::html_vignette
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{cgd_analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(purrr)
library(reshape2)
library(dplyr)
library(rstan)
library(dplyr)
library(glue)
library(abind)
library(survival)
library(multimcm)
library(frailtyHL)

library(multimcm)

# rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores() - 1)
```

# Introduction

The analysis is a comparison with a frequentist analysis in @Lai2009.


# Data

The data is chronic granulomatous disease (CGD) data available as part of the `{frailtyHL}` package.

```{r}
data(cgd)
head(cgd)
```

Create a new variable for the duration until the event.
```{r}
cgd <- mutate(cgd, time = tstop - tstart)
# center_id = as.numeric(as.factor(center))
```

The data has repeat measurements and 13 hospitals so for simplicity we'll remove some of the data in the initial analyses.
Take the first event for each individual by start time.
```{r first-measurement}
cgd_first <-
  cgd |> 
  group_by(id) |> 
  arrange(tstart) |> 
  filter(row_number() == 1)
```

Drop any levels which only appeared in the data we've just removed and convert the variables to numeric or factor for unique and categorical data, respectively.

```{r filter-center}
cgd_center <- cgd_first |>
  filter(center %in% c("Amsterdam", "NIH", "Univ. of Zurich", "Scripps Institute")) |>
  droplevels() |> 
  mutate(center_id = as.numeric(center),
         hos.cat_id = as.numeric(hos.cat),
         sex_id = as.factor(as.numeric(sex)))
```

Next, append the background hazard rate by age, sex and country.
We first need to harmonise the country names and their associated cities.
Then we join the data with the background mortality data according to age, sex and country.
```{r bg-hazard}
cgd_center <-
  cgd_center |>
  tidyr::separate(hos.cat, c("country", "hospital"),
                  remove = FALSE) |> 
  mutate(country = toupper(country),
         age_event = round(age + time/365.25, 0),   # convert from days to years
         country = ifelse(center == "Amsterdam", "NETHERLANDS",
                          ifelse(center == "Univ. of Zurich", "SWITZERLAND",
                                 ifelse(center == "Copenhagen", "DENMARK", country)))) |> 
  select(-hospital)

# background mortality data
load(here::here("../bgfscure/data/bg.mortality.RData"))

# haramonise fields
bg.mortality <- bg.mortality |> 
  mutate(ACOUNTRY = toupper(ACOUNTRY),
         ACOUNTRY = ifelse(ACOUNTRY == "UNITED STATES", "US", ACOUNTRY),
         SEX = ifelse(SEX == "M", "male", "female"),
         rate = ifelse(rate == 0, 1e-10, rate)) |>  # replace so >0
  rename(country = ACOUNTRY,
         sex = SEX,
         age_event = Age)

input_data <- merge(cgd_center, bg.mortality,
                    by = c("age_event", "sex", "country"), sort = FALSE)
```

That is the last step in the data wrangling. We can now proceed to the analysis.

# Analysis

### By center

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda)\\
\pi_i = \text{logit}^{-1}(\alpha + \beta_{\text{treat}[i]} + \gamma_{\text{center}[i]})\\
\text{center}[i] \sim N(\mu_{\text{center}}, \sigma^2_{\text{center}})
$$

```{r}
out <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400)
```

Alternatively, we can pass in a _precompiled_ model, rather than have the model compiled at the time of fitting.

```{r}
model_nm <- "precompiled_stan_model"

# create the precompiled model and save as .RDS file
precompile_bmcm_model(input_data = input_data,
                      cureformula =  "~ treat + (1 | center_id)",
                      family_latent = "exponential",
                      model_name = model_nm)
```

Now, load back in the compiled model and fit as before.

```{r}
model_path <- glue::glue("{system.file('stan', package = 'multimcm')}/{model_nm}.RDS")

out_precompiled <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path)
```

We can do the same but using a `cmdstanr` model, rather than an `rstan` model. This has some benefits in terms of speed, reliability and memory usage.
Firstly for a non-compiled model.

```{r}
out <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    use_cmdstanr = TRUE)

# so that plotting functions work later
# convert cmdstanr object to rstan stanfit format

##TODO: error. for some reason save_warmup is NA?...
# out$output <- rstan::read_stan_csv(out$output$output_files())
```

Secondly, we can pass in a _precompiled_ `cmdstanr` model. This saves an `.exe` file in the `stan` folder.

```{r}
model_nm <- "precompiled_stan_model"

precompile_bmcm_model(input_data = input_data,
                      cureformula =  "~ treat + (1 | center_id)",
                      family_latent = "exponential",
                      model_name = model_nm,
                      use_cmdstanr = TRUE)
```

and load back in the compiled model and fit as before.

```{r}
model_path <- glue::glue("{system.file('stan', package = 'multimcm')}/{model_nm}.exe")

out_precompiled <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path,
    use_cmdstanr = TRUE)

##TODO:
# out_precompiled$output <-
#   rstan::read_stan_csv(out_precompiled$output$output_files())
```

#### Separate models

We can also fit the end-points separately by changing the `cureformula` argument to a formula with `~ treat + center_id`.

```{r}
stan_model <- 
  precompile_bmcm_model(input_data = input_data,
                        cureformula =  "~ treat + center_id",
                        family_latent = "exponential",
                        model_name = "precompiled_stan_model",
                        use_cmdstanr = TRUE)

model_path <- stan_model$exe_file()

out_precompiled_separate <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + center_id",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400,
    precompiled_model_path = model_path,
    use_cmdstanr = TRUE)
```

### By hospital category

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda)\\
\pi_i = \text{logit}^{-1}(\alpha + \beta_{\text{treat}[i]} + \gamma_{\text{cat}[i]})\\
\text{cat}[i] \sim N(\mu_{\text{cat}}, \sigma^2_{\text{cat}})
$$

```{r}
out_hos.cat <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | hos.cat_id)",
    family_latent = "exponential",
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 365)
```


### With sex covariate in latent model

No covariates in the latent model and the incidence (cure) model as
$$
T \sim \text{Exp}(\lambda_i)\\
\lambda_i = \alpha_{\lambda} + \beta_{\text{sex}[i]}\\
\pi_i = \text{logit}^{-1}(\alpha_{\pi} + \beta_{\text{treat}[i]} + \gamma_{\text{center}[i]})\\
\text{center}[i] \sim N(\mu_{\text{center}}tyg, \sigma^2_{\text{center}})
$$

```{r}
out_sex <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ sex",
    cureformula = "~ treat + (1 | hos.cat_id)",
    family_latent = "exponential",
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 365)
```


## Plots

After fitting the models, we can plot the survival curves for each.

```{r}
library(ggplot2)
```

### Main models

```{r}
gg <- plot_S_joint(out,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
gg <- plot_S_joint(out_precompiled,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
gg <- plot_S_joint(out_precompiled_separate,
                   add_km = TRUE,
                   annot_cf = FALSE)
gg + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg, filename = "plots/cgd_surv_plot.png", dpi = 640, width = 10, height = 8)
```

### By center

```{r}
gg2 <- plot_S_joint(out_hos.cat,
                    add_km = TRUE,
                    annot_cf = FALSE)
gg2 + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg2, filename = "plots/cgd_hospcat_surv_plot.png", dpi = 640, width = 10, height = 8)
```

```{r}
gg3 <- plot_S_joint(out_sex,
                    add_km = TRUE,
                    annot_cf = FALSE)
gg3 + xlim(0,365) + facet_wrap(~endpoint)
```

```{r eval=FALSE, echo=FALSE}
ggsave(plot = gg3, filename = "plots/cgd_sex_surv_plot.png", dpi = 640, width = 10, height = 8)
```


## References
