---
title: "How to use multimcm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use multimcm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Introduction

This is a quick start, basic introduction to how to use `multimcm` to fit Bayesian mixture cure models in Stan.

## Data

We will use the xxx data set.
The data have already been arranged in to the correct format and saved within the package so we can load it as follows.

```{r eval=FALSE}
data("surv_input_data", package = "multimcm")
```

If you want to load the data set directly you can use
```{r echo=FALSE}
load("~/R/multimcm/data/surv_input_data.RData")
```

Required fields include event times (`os`, `pfs`) and censoring indicators (`os_event`, `pfs_event`) for both OS and PFS.
There should also be a treatment label column (`TRTA`).
Additional patient-level covariates can also be included.
At present only age at event (`OSage`, `PFSage`) is used.

This looks like this.

```{r}
head(surv_input_data)
```

## Example

First of all, attach all of the libraries we are going to need.

```{r message=FALSE, warning=FALSE}
library(purrr)
library(reshape2)
library(dplyr)
library(rstan)
library(shinystan)
library(dplyr)
library(ggplot2)
# library(multimcm)
```

For demonstration purposes we will select a single treatment and fit Exponential distributions to both OS and PFS.

```{r}
i <-  "exp"
k <- "exp"
```

To use the Stan engine we set some options to use all-but-one of the available cores and not to over-write pre-complied code.

```{r}
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores() - 1)
```

Now we are ready to do the model fitting using `bmcm_stan()`. Key arguments are `formula` and `cureformula`. These define the two levels observed (incidence) and latent of the model, respectively. They employ the standard `R` formula syntax; For `formula` using the survival regression syntax and mixed effect syntax for `cureformula`. Other arguments can be set to their default values.

```{r eval=FALSE}
out <-
  bmcm_stan(
    input_data = input_data,
    formula = "Surv(time=time, event=status) ~ 1",
    cureformula = "~ treat + (1 | center_id)",
    family_latent = "exponential",
    # prior_cure = NA,
    centre_coefs = TRUE,
    bg_model = "bg_fixed",
    bg_varname = "rate",
    bg_hr = 1,
    t_max = 400)
```


The `bg_model` is the background model. The current options are Exponential distribution (1) or empirical point estimates from WHO life-tables (2). If the former then the function uses a default prior. The remaining arguments are passed directly to the Stan engine for the HMC.

-   Finally, the remaining arguments are passed directly to the Stan engine for the HMC.

## Results

There are several very good packages available to view Stan output, including `shinystan` and `coda`. Below we give some basic output specific to `multimcm`.

### Output data

```{r echo=FALSE}
```

We can view the raw output from running the Stan model using

```{r}
res <- extract(stan_exp_exp)
```

The available posterior samples are

```{r}
names(res)
```

-   Parameters prefixed with `p` are prior predicted distribution samples.
-   Parameters prefixed with `lp_` (except `lp__`) are the linear predictors used for the Exponential distributions rate parameters `lambda_`.

### Plots

The function `plot_S_joint()` takes a list of multiple Stan runs and creates `ggplot2` grid object of survival curves for cured, uncured and mixed with 95% credible intervals.
In our example we simply create a list of one element.

```{r warning=FALSE, message=FALSE}
stan_list <- list("IPILIMUMAB" = stan_exp_exp_IPI)
gg <- plot_S_joint(stan_list = stan_list)
```

In addition, we overlay the Kaplan-Meier curves for the original data using the `survival` package.
Note that this will always be slightly different since this is for the trial case-mix and the posterior survival curves are for the *average* patient.

```{r fig.width=7, warning=FALSE, message=FALSE}
library(survival)

fit_os <- survfit(Surv(os, os_event) ~ 1,
                  data = filter(surv_input_data, TRTA == trta))
fit_pfs <- survfit(Surv(pfs, pfs_event) ~ 1,
                   data = filter(surv_input_data, TRTA == trta))
km_data <-
  rbind(
    data.frame(Tx = trta,
               event_type = "os",
               time = fit_os$time,
               surv = fit_os$surv),
    data.frame(Tx = trta,
               event_type = "pfs",
               time = fit_pfs$time,
               surv = fit_pfs$surv))

gg + geom_line(aes(x = time, y = surv),
               data = km_data,
               lwd = 1,
               inherit.aes = FALSE) +
  xlim(0, 60) + ylim(0, 1)
```
